---
title: "Statistical analysis - based on household surveys"
author: "Franziska Walther"
date: "03/10/2023"
output: html_document
---

#### (Install and) load packages:     
```{r,warning=FALSE,message=FALSE}
setwd("~/Pius_stats")

library(rstan)
library(brms)
library(readxl)
library(tidyverse)
library(bayesplot)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

### Research hypothesis 1:  
Drivers for crop diversification -    
How does crop diversification vary across households?   
The relative contribution of household measures (predictors) to dominant adaptation (i.e. crop diversification, reponse).    

##### Data wrangling  
```{r}
#load data 
data <- read_xlsx("~/Pius_stats/data_indicators_new.xlsx")
data_crop_div <- read_xlsx(("~/Pius_stats/data_hhperc_crops.xlsx"))

# reduce multiple instances of one household
crop_div_reduced <-  data_crop_div %>% 
  group_by(hh_index) %>% 
  summarise(crop_diversification = sum(`Crop diversification`))
  
# merge data sets
data_merge <- merge(data, crop_div_reduced, by.x = "hh_index", by.y = "hh_index", all.x = TRUE)

# handle multiple crop diversifications in one household and NAs
data_merge$crop_diversification <- ifelse(is.na(data_merge$crop_diversification), 0,
                                                data_merge$crop_diversification)
data_merge$crop_diversification <- ifelse(data_merge$crop_diversification > 1, 1,
                                          data_merge$crop_diversification)
data_merge <- data_merge %>% 
  mutate_if(is.character,as.factor)

#Create dataframe of interest (response and predictors)
df_m <- data_merge %>%
  select(crop_diversification, hhh_gender, hhh_age, hhsize, hh_landsize, hh_plots, hh_farmyears, assets_number_hh, income_annual_usd, income_sources_number, liv_tlu_total, credit, use_any, hhh_educyears)

#Some variables need to be turned into factors
df_m$hhh_gender <- as.factor(df_m$hhh_gender)
df_m$credit <- as.factor(df_m$credit)
df_m$use_any <- as.factor(df_m$use_any)

str(df_m) 

```

##### Modelling:    
The response variable is binary (0 or 1), hence I decided for a logistic regression using a binomial distribution, i.e. bernoulli distribution. 
```{r}

fit <- brm(crop_diversification ~ hhh_gender + hhh_age + hhsize + hh_landsize + hh_plots + hh_farmyears + assets_number_hh + income_annual_usd + income_sources_number + liv_tlu_total + credit + use_any + hhh_educyears, data=df_m, family = bernoulli, chains=4, iter = 2000, warmup=1000)

```

##### Have a look at model diagnostics and outcome:   
```{r}
rhats <- rhat(fit)
mcmc_rhat(rhats)
ratios <- neff_ratio(fit)
mcmc_neff(ratios)

#Posterior predictive check 
pp_check(fit)

prior_summary(fit)
```

##### Specify priors:  
Uniform/Flat priors are the least informative assumptions, hence I adjusted the priors.   
```{r}
fit1 <- brm(crop_diversification ~ hhh_gender + hhh_age + hhsize + hh_landsize + hh_plots + hh_farmyears + assets_number_hh + income_annual_usd + income_sources_number + liv_tlu_total + credit + use_any + hhh_educyears, 
             data=df_m, family = bernoulli, chains=4, iter = 2000, warmup=1000, 
             prior = prior(normal(0,10), class = b))
            
```

##### Have a look at model outcome:   
```{r}
#PRIOR predictive distribution 
  m1_prior <- brm(crop_diversification ~ hhh_gender + hhh_age + hhsize + hh_landsize + hh_plots + hh_farmyears + assets_number_hh + income_annual_usd + income_sources_number + liv_tlu_total + credit + use_any + hhh_educyears, 
             data=df_m, family = bernoulli, chains=4, iter = 2000, warmup=1000, 
             prior = prior(normal(0, 10), class = b),
            sample_prior = "only")

  #prior predictive checks
  pp_check(m1_prior)

#POSTERIOR
posterior_summary(fit1)

#posterior predictive checks 
pp_check(fit1)

#Slopes: Predictors' effects on response  
mcmc_plot(fit1, point_est = "median", prob = .6, prob_outer = .90)

#potential drivers 
mcmc_intervals(fit1, point_est = "median", pars = c("b_hhh_genderMale", "b_creditYes", "b_liv_tlu_total","b_use_any1","b_hhh_educyears"),prob=0.9)

#model diagnostics 
  #chains: 
  mcmc_plot(fit1, type = "trace")
   rhats <- rhat(fit1)
  mcmc_rhat(rhats)
  ratios <- neff_ratio(fit1)
  mcmc_neff(ratios)

```

##### Preliminary findings:   
**Crop diversification is affected by:**  
**- gender (male: 2.11 [0.06-4.56]),**     
**- use of weather information (yes: 3.05 [1.07-5.35]),**     
**- credit -1.04[-2.22 - 0.17]),**     
**- years of education (0.17 [0.04 - 0.31),**     
**- (and livestock TLU (0.11 [-0.04 - 0.28])).**     
**- (median [95% confidence interval])**    


### Research hypothesis 2:   
How does the use of weather information vary across households?   

#### Prepare data and create Model:  
```{r}
data <- read_xlsx("~/Pius_stats/data_indicators_new.xlsx")

data_2 <- data %>%
  select(use_any, access_number, income_annual_usd, income_sources_number, hhh_gender, hhh_age, hhh_educyears, hh_farmyears, livestock_have)

data_2$use_any <- as.factor(data_2$use_any)
data_2$hhh_gender <- as.factor(data_2$hhh_gender)
data_2$livestock_have <- as.factor(data_2$livestock_have)

str(data_2) 

fit2 <- brm(use_any ~ income_annual_usd + income_sources_number + hhh_gender + hhh_age +
              hhh_educyears + hh_farmyears + livestock_have, 
             data=data_2, family = bernoulli, chains=4, iter = 2000, warmup=1000, 
            prior = c(prior(normal(0,10), class=b, coef="income_annual_usd"), 
            #how to include non-normal distribution of income (e.g. exp)? 
                       prior(normal(0,5), class=b, coef="income_sources_number"), 
                       prior(normal(0,5), class = b, coef = "hhh_educyears"),
                       prior(normal(0,5), class = b, coef = "hh_farmyears")))
            #how to set priors for categorical variables!? 
            # prior(normal(0,1), class = b, coef = "livestock_have1"),
            # prior(normal(0,1), class = b, coef = "hhh_genderMale")))
```


##### Have a look at model diagnostics and outcome:   
```{r}
#model diagnostics 
rhats <- rhat(fit2)
mcmc_rhat(rhats)

ratios <- neff_ratio(fit2)
mcmc_neff(ratios)

#chains: 
  mcmc_plot(fit2, type = "trace")

#POSTERIOR
#posterior predictive checks 
pp_check(fit2)

posterior_summary(fit2)

#potential drivers:
mcmc_intervals(fit2, point_est = "median", pars = c("b_income_sources_number", "b_hhh_genderMale", "b_hhh_age", "b_hh_farmyears","b_livestock_have1"),prob=0.9)

```

##### Preliminary findings:   
**The use of weather information is affected by households':**  
**- income source number (0.47 [-0.04 - 1.03])(?),**     
**- gender (male: 0.38 [-0.6 - 1.28])(?),**     
**- age (0.10 [0.06 - 0.15]),**   
**- years of farming (-0.07 [ -0.11 - -0.02]),**     
**- and if the household has livestock (yes: -0.42 [-1.18 - 0.32])(?).**     
**- ? = significant or not, bc 95% CI transgresses 0.**     


### Research hypothesis 3:   

How does income vary across households? Which factors are driving income disparities?   
```{r}
data <- read_xlsx("~/Pius_stats/data_indicators_new.xlsx")

data_3 <- data %>%
  select(income_annual_usd, income_sources_number, hhh_gender, hhh_age, assets_number_hh, liv_tlu_total, hhh_educyears, hh_farmyears, credit, hhsize, hh_landsize, hh_plots)

data_3$hhh_gender <- as.factor(data_3$hhh_gender)
data_3$credit <- as.factor(data_3$credit)
data_3$income_annual_usd <- data_3$income_annual_usd + 0.01 
str(data_3) 

#response: annual income in USD
fit3A <- brm(income_annual_usd ~ hhh_gender + hhh_age + assets_number_hh + liv_tlu_total + hhh_educyears + hh_farmyears + credit + hhsize + hh_landsize + hh_plots, 
             data=data_3, family = exponential,
            chains=4, iter = 2000, warmup=1000, 
             prior = prior(normal(0,5), class=b))
                    
#response: number of income sources 
fit3B <- brm(income_sources_number ~ hhh_gender + hhh_age + assets_number_hh + liv_tlu_total + hhh_educyears + hh_farmyears + credit + hhsize + hh_landsize + hh_plots, 
             data=data_3, family = poisson, 
            chains=4, iter = 2000, warmup=1000, 
             prior = prior(normal(0,5), class=b))
```


##### Have a look at model 3A:   
```{r}
  #model diagnostics
  rhats <- rhat(fit3A)
  mcmc_rhat(rhats)
  ratios <- neff_ratio(fit3A)
  mcmc_neff(ratios)
  #chains: 
  mcmc_plot(fit3A, type = "trace")

#POSTERIOR
#posterior predictive checks 
pp_check(fit3A)

posterior_summary(fit3A)

#potential drivers:
mcmc_intervals(fit3A, point_est = "median", pars = c("b_hhh_genderMale", "b_hhh_age", "b_assets_number_hh","b_liv_tlu_total", "b_hh_farmyears", "b_creditYes", "b_hhsize", "b_hh_landsize", "b_hh_plots"),prob=0.9)

```

**Model 3A indicates that gender, age, number of assets, TLU, years of farming, credit, household size, landsize, plots effect the annual income (USD).**   


##### Have a look at model 3B:   
```{r}
  #Model diagnostics 
  rhats <- rhat(fit3B)
  mcmc_rhat(rhats)
  ratios <- neff_ratio(fit3B)
  mcmc_neff(ratios)
  #chains: 
  mcmc_plot(fit3B, type = "trace")

  #potential drivers: 
mcmc_intervals(fit3B, point_est = "median", pars = c("b_assets_number_hh","b_liv_tlu_total", "b_hh_landsize"),prob=0.9)
```

**Model 3B indicates that the number of income sources is influenced by the number of assets, TLU and the landsize of households.**    